Xtensible Binary Format (XBF): An Efficient Self-Describing Binary Format

### 2023-08-20 docs/paper/paper.tex
Update presentation and paper

% An additional optimization that could be made (but isn't with either of these formats) is deduplication of the metadata. Both JSON and XML include the metadata for every element sent, even if what is being sent is a homogenous list of elements that all share the same structure.
%

@@

Strings should always be the final value in the list. The value given to strings
is used by Vectors and Structs to determine what their discriminant value should
be.

@@

Vectors are a homogenous list of values that has a known, variable length. Vectors should first include their length as an unsigned 64-bit integer (the same as Strings and Bytes), followed by the corresponding number of elements. The type contained within a Vector is not sent to the client. That information is carried in the metadata.

@@

Similarly to primitives, a single byte discriminant value signifying a Vector is incoming is sent. This discriminant value should be 1 greater than that of the discriminant value for Strings. Following this, metadata information for the internal type contained within the Vector will be sent. This process may continue recursively with nested types of Vectors and Structs. The length of a vector or the data contained within the vector must not be sent.

@@

A Struct is an aggregate type containing a name as well as named fields. A struct may not contain duplicate field names. Should a Struct be sent like this anyway, it should be considered malformed and not be constructed on the receiving end. Fields of a Struct are sent in sequence in the order they are listed in the Struct's metadata. When a struct is serialized it should not send any sort of name information (such as its name or field names), how many fields it has, nor should it send any type information about its fields. That information is carried in the metadata.

@@

A discriminant value should first be sent, similarly to primitives (following the same size requirement). This discriminant value should be 1 greater than that of the discriminant value for Vectors.

Following this, the name of the Struct should be sent, using the same format as primitive strings are sent (unsigned 64-bit length and then the bytes). Next, send the number of fields contained within the Struct as an unsigned 16-bit integer. A 16-bit integer is used because it's very unlikely for a struct of more than 65 thousand fields to be necessary. Finally, the fields of the Struct should be sent, first the name of the field as a String, then immediately after the metadata for the type of the field. This process may continue recursively with nested types of Structs or Vectors. These name and type pairs will be sent until there are no more fields left in the Struct.

@@

The struct has a name Point, and two fields, named first and last, both of type i8. The goal is to serialize the variable \texttt{my\_point}, which is an instance of Point with the values 42 and -5 for the members first and last respectively. The struct's metadata should be sent as follows:

@@

The values inside the struct should then be sent in the same order as they are listed in the struct's metadata:

@@

\begin{table}[htbp]
	\caption{XBF Type Map}\label{type_map}
	\begin{center}
		\begin{tabular}{|c|c|c|}
			\hline
			\textbf{XBF Type} & \textbf{Rust Type}          & \textbf{C++ Type}                         \\
			\hline
			Boolean           & bool                        & bool                                      \\
			U8                & u8                          & uint8\_t                                  \\
			U16               & u16                         & uint16\_t                                 \\
			U32               & u32                         & uint32\_t                                 \\
			U64               & u64                         & uint64\_t                                 \\
			U128              & u128                        & uint64\_t[2]                              \\
			U256              & [u64; 4]                    & uint64\_t[4]                              \\
			I8                & i8                          & int8\_t                                   \\
			I16               & i16                         & int16\_t                                  \\
			I32               & i32                         & int32\_t                                  \\
			I64               & i64                         & int64\_t                                  \\
			I128              & i128                        & uint64\_t[2]                              \\
			I256              & [u64; 4]                    & uint64\_t[4]                              \\
			F32               & f32                         & float                                     \\
			F64               & f64                         & double                                    \\
			Bytes             & Vec\textless u8\textgreater & std::vector\textless uint8\_t\textgreater \\
			String            & String                      & std::string                               \\
			\hline
		\end{tabular}
	\end{center}
\end{table}

@@

In Rust there is an important distinction between an owned String, and a string slice\cite{strings}. A string slice is similar to the previously mentioned slices, except it consists of valid UTF-8 code points. It is a view into a block of memory, and its size cannot be modified. If growable string is required, that's where the String type is used. It makes the same UTF-8 guarantee as a string slice, but has the same properties as a Vec where it keeps track of its capacity, and will reallocate if its capacity is exceeded, similar to a dynamic array. As with the decision to use a boxed slice in XbfStruct, a boxed string slice is utilized to save on memory usage, as the metadata for a struct shouldn't need to change, thereby making a capacity field irrelevant.

@@

>>>>

@@

Strings will always be the final value in the list. The value given to strings
is used by Vectors and Structs to determine what their discriminant value will
be.

@@

Vectors are a homogenous list of values that has a known, variable length. Vectors will first include their length as an unsigned 64-bit integer (the same as Strings and Bytes), followed by the corresponding number of elements. The type contained within a Vector is not sent to the client. That information is carried in the metadata.

@@

Similarly to primitives, a single byte discriminant value signifying a Vector is incoming is sent. This discriminant value will be 1 greater than that of the discriminant value for Strings. Following this, metadata information for the internal type contained within the Vector will be sent. This process may continue recursively with nested types of Vectors and Structs. The length of a vector or the data contained within the vector must not be sent.

@@

A Struct is an aggregate type containing a name as well as named fields. A struct may not contain duplicate field names. Should a Struct be sent like this anyway, it is considered malformed and may not be constructed on the receiving end. Fields of a Struct are sent in sequence in the order they are listed in the Struct's metadata. When a struct is serialized it does not send any sort of name information (such as its name or field names), how many fields it has, nor does it send any type information about its fields. That information is carried in the metadata.

@@

A discriminant value will first be sent, similarly to primitives (following the same size requirement). This discriminant value is 1 greater than that of the discriminant value for Vectors.

Following this, the name of the Struct will be sent, using the same format as primitive strings are sent (unsigned 64-bit length and then the bytes). Next, send the number of fields contained within the Struct as an unsigned 16-bit integer. A 16-bit integer is used because it's very unlikely for a struct of more than 65 thousand fields to be necessary. Finally, the fields of the Struct will be sent, first the name of the field as a String, then immediately after the metadata for the type of the field. This process may continue recursively with nested types of Structs or Vectors. These name and type pairs will be sent until there are no more fields left in the Struct.

@@

The struct has a name Point, and two fields, named first and last, both of type i8. The goal is to serialize the variable \texttt{my\_point}, which is an instance of Point with the values 42 and -5 for the members first and last respectively. The struct's metadata will be sent as follows:

@@

The values inside the struct will be sent in the same order as they are listed in the struct's metadata:

@@

\begin{table}[htbp]
	\caption{XBF Type Map}\label{type_map}
	\begin{center}
		\begin{tabular}{|c|c|c|}
			\hline
			\textbf{XBF Type} & \textbf{Rust Type}          & \textbf{C++ Equivalent Type}              \\
			\hline
			Boolean           & bool                        & bool                                      \\
			U8                & u8                          & uint8\_t                                  \\
			U16               & u16                         & uint16\_t                                 \\
			U32               & u32                         & uint32\_t                                 \\
			U64               & u64                         & uint64\_t                                 \\
			U128              & u128                        & uint64\_t[2]                              \\
			U256              & [u64; 4]                    & uint64\_t[4]                              \\
			I8                & i8                          & int8\_t                                   \\
			I16               & i16                         & int16\_t                                  \\
			I32               & i32                         & int32\_t                                  \\
			I64               & i64                         & int64\_t                                  \\
			I128              & i128                        & uint64\_t[2]                              \\
			I256              & [u64; 4]                    & uint64\_t[4]                              \\
			F32               & f32                         & float                                     \\
			F64               & f64                         & double                                    \\
			Bytes             & Vec\textless u8\textgreater & std::vector\textless uint8\_t\textgreater \\
			String            & String                      & std::string                               \\
			\hline
		\end{tabular}
	\end{center}
\end{table}

@@

In Rust there is an important distinction between an owned String, and a string slice\cite{strings}. A string slice is similar to the previously mentioned slices, except it consists of valid UTF-8 code points. It is a view into a block of memory, and its size cannot be modified. If growable string is required, that's where the String type is used. It makes the same UTF-8 guarantee as a string slice, but has the same properties as a Vec where it keeps track of its capacity, and will reallocate if its capacity is exceeded, similar to a dynamic array. As with the decision to use a boxed slice in XbfStruct, a boxed string slice is utilized to save on memory usage, as the metadata for a struct won't need to change, thereby making a capacity field irrelevant.

@@

See Table \ref{time_and_bytes} and \ref{overhead} for additional results. Overhead is calculated as native data size recorded by the server subtracted from bytes sent for a format.

### 2023-08-20 docs/presentation/presentation.tex

\title{Implementing XBF: An Efficient Self-Describing Binary Format}
\author{\href{mailto:dkrautha@pm.me}{David Krauthamer}}

@@

\begin{frame}{Web Data Interchange Formats}
	\begin{itemize}
		\item Human-Readable and Self-Describing
		      \note[item]{Most formats on the web are human-readable and self-describing}
		      \note[item]{Self describing means they include metadata to describe themselves, and don't require an external schema}
		      \note[item]{Human readable means plain text, ASCII or utf-8}
		\item Most common: JSON and XML
		      \note[item]{The most common of these formats is JSON and XML}
		\item Alternatives: CSV, MessagePack, CBOR
		      \note[item]{There are other human-readable formats such as CSV}
		      \note[item]{And others, such as CBOR (concise binary object representation), and MessagePack that are still self-describing, but use a binary format instead}
	\end{itemize}
\end{frame}

\begin{frame}{Shortcomings of Current Formats}
	\begin{itemize}
		\item Repeated sending of metadata.
		\item Inefficiency of plain text.
		      \note[item]{Expensive conversion between plain text and binary}
		      \note[item]{Requires more bytes to store the same data}
		\item Use of big-endian.
		      \note[item]{Most modern processors (x86 and ARM) are little-endian}
		      \note[item]{Both CBOR and MessagePack are big-endian}
	\end{itemize}
	% XML from paper here
\end{frame}

\begin{frame}[fragile]{Repeated Metadata}
	\begin{columns}
		\begin{column}{0.5\textwidth}
			\note[item]{XML example where a huge percent of the actual data sent is metadata}
			\begin{lstlisting}[language=XML]
<person>
	<name>John Jackson</name>
	<age>25</age>
</person>
\end{lstlisting}
		\end{column}
		\begin{column}{0.5\textwidth}
			\begin{table}
				\caption{XML Space Utilization}
				\center
				\begin{tabular}{|c|c|c|}
					\hline
					\textbf{} & \textbf{Bytes Used} & \textbf{Percent} \\
					\hline
					Metadata  & 45                  & 76\%             \\
					Data      & 14                  & 24\%             \\
					\hline
					Total     & 59                  & 100              \\
					\hline
				\end{tabular}
			\end{table}
		\end{column}
	\end{columns}
\end{frame}

@@

@@

\begin{frame}{Improvements For a New Format}
	\begin{itemize}
		\item When possible, send metadata only a single time (mainly for homogenous lists).
		\item Allow sending metadata to be optional, such as if a client already has it.
		      \note[item]{Keep metadata and data separate, not coupled together}
		\item Utilize little-endian to eliminate endian conversions on common architectures.
		\item Prioritize a simple type system.
		      \note[item]{Hopefully makes the implementation simpler}
	\end{itemize}
\end{frame}

@@

\begin{frame}{Primitives and Primitive Metadata}
	\begin{columns}
		\begin{column}{0.5\textwidth}
			\begin{itemize}
				\item Boolean
				      \note[item]{True of False, one byte}
				      \note[item]{All integers are little-endian}
				      \note[item]{Integer types are equivalent to the types in stdint.h C headers}
				\item U8, U16, U32, U64, U128, U256
				      \note[item]{Unsigned Integer Types}
				      \begin{itemize}
					      \item Equivalent to uint*\_t types in C
				      \end{itemize}
				\item I8, I16, I32, I64, I128, I256
				      \note[item]{Signed Integer Types}
				      \begin{itemize}
					      \item Equivalent to int*\_t types in C
				      \end{itemize}
				\item F32, F64
				      \note[item]{Floating Point Types}
				      \begin{itemize}
					      \item IEEE 754 floating point numbers
					      \item Equivalent to float and double in C
				      \end{itemize}
				\item Bytes  (Non-UTF-8 Bytes)
				      \note[item]{String but doesn't have to be UTF-8}
				\item String (UTF-8 Bytes)
				      \note[item]{Sends length as u64 followed by UTF-8 bytes}
			\end{itemize}
		\end{column}
		\begin{column}{0.25\textwidth}
			\center
			\begin{tabular}{|c|c|}
				\hline
				\textbf{Type} & \textbf{Value} \\
				\hline
				Boolean       & 0              \\
				U8            & 1              \\
				U16           & 2              \\
				U32           & 3              \\
				U64           & 4              \\
				U128          & 5              \\
				U256          & 6              \\
				\hline
			\end{tabular}
		\end{column}
		\begin{column}{0.25\textwidth}
			\center
			\begin{tabular}{|c|c|}
				\hline
				\textbf{Type} & \textbf{Value} \\
				\hline
				I8            & 7              \\
				I16           & 8              \\
				I32           & 9              \\
				I64           & 10             \\
				I128          & 11             \\
				I256          & 12             \\
				F32           & 13             \\
				F64           & 14             \\
				Bytes         & 15             \\
				String        & 16             \\
				\hline
			\end{tabular}
		\end{column}
	\end{columns}
\end{frame}

@@

\begin{frame}{Struct Metadata}
	\begin{itemize}
		\item Single byte discriminant value is sent (Vector + 1).
		\item Followed by Struct name.
		\item Next, number of fields as an unsigned 16-bit integer.
		      \note[item]{A 16-bit integer is used because it's very unlikely for a struct of more than 65 thousand fields to be necessary.}
		\item Last, pairs of field names and types.
		      \begin{itemize}
			      \item May continue recursively with nested types of Structs or Vectors.
		      \end{itemize}
		\item All names are sent with the same format as primitive Strings.
	\end{itemize}
\end{frame}

@@

\begin{frame}{Encoder and Decoder Reference Implementation}
	\begin{itemize}
		\item Written in Rust.
		      \note[item]{systems level programming language similar to C and C++, strong memory safety guarantees}
		\item Extensively Tested.
		      \note[item]{100 percent code coverage}
		      \note[item]{not going to talk a ton about the reference implementation}
		\item Majority of time was spent on this implementation.
	\end{itemize}
\end{frame}

\begin{frame}{Performance Test}
	\begin{columns}
		\begin{column}{0.5\textwidth}
			Server
			\begin{itemize}
				\item Downloads and parses 1 year of Sony stock history.
				\item 
			\end{itemize}
		\end{column}
		\begin{column}{0.5\textwidth}
				\item For each data format the client:
				      \begin{itemize}
					      \item Client sends 100 requests to the server.
					      \item Records average time to receive a response.
					      \item Records number of bytes read.
				      \end{itemize}
		\end{column}
	\end{columns}
\end{frame}

\section{Results}

\begin{frame}{Average Time and Bytes Read}
	\center
	\begin{tabular}{|c|c|c|c|c|}
		\hline
		\textbf{Format} & \textbf{Avg Time (ms)} & \textbf{Bytes Read} & \textbf{Overhead (bytes)} & \textbf{Percent Overhead} \\ 
		\hline                                                                                                                    
		CSV             & 18.93                  & 16,411              & 1,823                     & 11.1                      \\ 
		MessagePack     & 11.22                  & 15,565              & 977                       & 6.22                      \\ 
		CBOR            & 16.95                  & 25,507              & 10,919                    & 42.8                      \\ 
		JSON            & 21.91                  & 31,180              & 16,592                    & 53.2                      \\ 
		XML             & 21.87                  & 43,699              & 29,111                    & 66.6                      \\ 
		XBF             & 11.32                  & 14,686              & 98                        & 0.67                      \\ 
		\hline
	\end{tabular}
\end{frame}

\begin{frame}{Overhead}
	\center
	\begin{tabular}{|c|c|c|}
		\hline
		\textbf{Format} & \textbf{Overhead (bytes)} & \textbf{Percent Overhead} \\
		\hline
		CSV             & 1,823                     & 11.1                      \\
		MessagePack     & 977                       & 6.22                      \\
		CBOR            & 10,919                    & 42.8                      \\
		JSON            & 16,592                    & 53.2                      \\
		XML             & 29,111                    & 66.6                      \\
		XBF             & 98                        & 0.67                      \\
		\hline
	\end{tabular}
\end{frame}

>>>>

\title{Xtensible Binary Format (XBF): An Efficient Self-Describing Binary Format}
\author{\href{mailto:dkrautha@pm.me}{David Krauthamer}}

@@

\begin{frame}{Web Data Interchange Formats}
	\begin{itemize}
		\item Most common: JSON and XML
		\item Human-Readable and Self-Describing
		\item Alternatives: CSV, MessagePack, CBOR
	\end{itemize}
\end{frame}

\begin{frame}{Shortcomings of Current Formats}
	\begin{itemize}
		\item Inefficiency of plain text.
		\item Repeated sending of metadata.
		\item Use of big-endian.
	\end{itemize}
\end{frame}

@@

\begin{frame}[fragile]{Repeated Metadata}
	\begin{columns}
		\begin{column}{0.5\textwidth}
			\begin{lstlisting}[language=XML]
<person>
	<name>John Jackson</name>
	<age>25</age>
</person>
\end{lstlisting}
		\end{column}
		\begin{column}{0.5\textwidth}
			\begin{table}
				\caption{XML Space Utilization}
				\center
				\begin{tabular}{|c|c|c|}
					\hline
					\textbf{} & \textbf{Bytes Used} & \textbf{Percent} \\
					\hline
					Metadata  & 45                  & 76\%             \\
					Data      & 14                  & 24\%             \\
					\hline
					Total     & 59                  & 100              \\
					\hline
				\end{tabular}
			\end{table}
		\end{column}
	\end{columns}
\end{frame}

@@

\begin{frame}{Improvements For A New Format}
	\begin{itemize}
		\item When possible, send metadata only a single time (mainly for homogenous lists).
		\item Allow sending metadata to be optional, such as if a client already has it.
		\item Utilize little-endian to eliminate endian conversions on common architectures.
		\item Prioritize a simple type system.
	\end{itemize}
\end{frame}

@@

\begin{frame}{Primitives and Primitive Metadata}
	\begin{columns}
		\begin{column}{0.5\textwidth}
			\begin{itemize}
				\item Boolean
				\item U8, U16, U32, U64, U128, U256
				      \begin{itemize}
					      \item Equivalent to uint*\_t types in C
				      \end{itemize}
				\item I8, I16, I32, I64, I128, I256
				      \begin{itemize}
					      \item Equivalent to int*\_t types in C
				      \end{itemize}
				\item F32, F64
				      \begin{itemize}
					      \item IEEE 754 floating point numbers
					      \item Equivalent to float and double in C
				      \end{itemize}
				\item Bytes  (Non-UTF-8 Bytes)
				\item String (UTF-8 Bytes)
			\end{itemize}
		\end{column}
		\begin{column}{0.25\textwidth}
			\center
			\begin{tabular}{|c|c|}
				\hline
				\textbf{Type} & \textbf{Value} \\
				\hline
				Boolean       & 0              \\
				U8            & 1              \\
				U16           & 2              \\
				U32           & 3              \\
				U64           & 4              \\
				U128          & 5              \\
				U256          & 6              \\
				\hline
			\end{tabular}
		\end{column}
		\begin{column}{0.25\textwidth}
			\center
			\begin{tabular}{|c|c|}
				\hline
				\textbf{Type} & \textbf{Value} \\
				\hline
				I8            & 7              \\
				I16           & 8              \\
				I32           & 9              \\
				I64           & 10             \\
				I128          & 11             \\
				I256          & 12             \\
				F32           & 13             \\
				F64           & 14             \\
				Bytes         & 15             \\
				String        & 16             \\
				\hline
			\end{tabular}
		\end{column}
	\end{columns}
\end{frame}

@@

\begin{frame}{Struct Metadata}
	\begin{itemize}
		\item Single byte discriminant value is sent (Vector + 1).
		\item Followed by Struct name.
		\item Next, number of fields as an unsigned 16-bit integer.
		\item Last, pairs of field names and types.
		      \begin{itemize}
			      \item May continue recursively with nested types of Structs or Vectors.
		      \end{itemize}
		\item All names are sent with the same format as primitive Strings.
	\end{itemize}
\end{frame}

@@

\begin{frame}{Encoder and Decoder Reference Implementation}
	\begin{itemize}
		\item Written in Rust
		      \begin{itemize}
			      \item Efficient code generation similar to C and C++
			      \item Built-in tagged unions
			      \item Strong memory safety guarantees
		      \end{itemize}
		\item Extensively tested
		      \begin{itemize}
			      \item Unit and integration tests
			      \item 100 percent code coverage
		      \end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Performance Test}
	\begin{columns}
		\begin{column}{0.5\textwidth}
			Multi-threaded Server
			\begin{itemize}
				\item Downloads and parses 1 year of Sony stock history.
				\item Records the original CSV file size in bytes.
				\item Records the size in bytes when stored natively in memory.
				\item Waits to receive requests.
			\end{itemize}
		\end{column}
		\begin{column}{0.5\textwidth}
			Client
			\begin{itemize}
				\item Formats: CSV, MessagePack, CBOR, JSON, XML, XBF
				\item For each data format:
				      \begin{itemize}
					      \item Client sends 100 requests to the server.
					      \item Records average time to receive a response.
					      \item Records number of bytes read.
				      \end{itemize}
			\end{itemize}
		\end{column}
	\end{columns}
\end{frame}

\section{Results and Conclusions}

\begin{frame}[fragile]{Data}
	\begin{center}
		\begin{tabular}{|c|c|c|c|c|}
			\hline
			\textbf{Format} & \textbf{Avg Time (ms)} & \textbf{Bytes Read} & \textbf{Overhead (bytes)} & \textbf{Percent Overhead} \\
			\hline
			CSV             & 18.93                  & 16,411              & 1,823                     & 11.1                      \\
			MessagePack     & 11.22                  & 15,565              & 977                       & 6.22                      \\
			CBOR            & 16.95                  & 25,507              & 10,919                    & 42.8                      \\
			JSON            & 21.91                  & 31,180              & 16,592                    & 53.2                      \\
			XML             & 21.87                  & 43,699              & 29,111                    & 66.6                      \\
			XBF             & 11.32                  & 14,686              & 98                        & 0.67                      \\
			\hline
		\end{tabular}
	\end{center}
	Original CSV data size recorded by the server: 17,160 bytes.

	Native data size recorded by the server: 14,558 bytes.
\end{frame}

\begin{frame}{CSV Size Difference}
	CSV Size Difference
	\begin{itemize}
		\item Original data always used 8 digits of precision.
		\item CSV library used only writes the required digits of precision.
		\item Example: 83.500000 becomes 83.5
	\end{itemize}
\end{frame}

\begin{frame}{Time Performance}
	\begin{itemize}
		\item XBF was faster than every format except MessagePack.
		\item Formats that operated in binary were faster, even when sending more data (CBOR).
		\item XBF reference implementation was optimized for memory usage, more could be done for speed.
	\end{itemize}
\end{frame}

\begin{frame}{Size Performance}
	\begin{itemize}
		\item XBF required the least amount of bytes.
		\item Indicates metadata deduplication worked as intended.
		\item XBF can be even smaller with optimization techniques from MessagePack
		      \begin{itemize}
			      \item Coercion to smaller integer sizes when possible.
			      \item Differently sized variable length types.
		      \end{itemize}
		\item Could be possible to have a "negative" overhead.
	\end{itemize}
\end{frame}

\begin{frame}{Use Cases}
	\begin{itemize}
		\item Internet of Things and other low bandwidth scenarios.
		\item High volume applications.
	\end{itemize}
\end{frame}
